Integrated nested Laplace approximations for extended latent Gaussian models

### 2022-11-12 src/docs_paper/paper.Rmd
More writing

---
title: Integrated nested Laplace approximations for extended latent Gaussian models
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler. By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - Spatial statistics
  - INLA

@@

The complexity of the model presents a difficult Bayesian inference problem.
Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
`TMB` is gaining popularity spatial statistics [@osgoodzimmerman2021statistical] owing to its speed and flexibility.
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
For the Naomi model, we use this option to integrate out the latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.

To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
INLA [@rue2009approximate] is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
The central innovation is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
This extension facilitates inference for models like Naomi, which fall within the extended latent Gaussian model (ELGM) [@stringer2022fast] class, and were not previously amenable to inference with INLA.
ELGMs build on latent Gaussian models (LGMs) by allowing each element of the linear predictor to depend on any subset of elements from the latent field.
We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
Though as of yet no theory exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
As `aghq` is designed to naturally interface with `TMB`, use of the method is easy when provided a \textsf{C++} user template for the log-posterior.

@@

Consider a country partitioned into $n$ areas indexed by $i$.
Suppose a simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
Cases may be modelled using a binomial logistic regression model
\begin{align}
y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
\end{align}
where HIV prevalence $\rho^\text{HS}_i$ is modelled by a Gaussian with mean $\beta_\phi$ and standard deviation $\sigma_\phi$.

@@

Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
Suppose that of $m^\text{ANC}_i$ women attending ANC, $y^\text{ANC}_i$ are HIV positive.
Then an analogous binomial logistic regression model
\begin{align}
y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
\end{align}
may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence on the logit scale.

@@

The number of people receiving treatment at district health facilities $A_i$ provides further information about HIV prevalence.
Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
ART coverage, defined to be the proportion of people living with HIV (PLHIV) currently on ART on district $i$, is given by $\alpha_i = A_i / \rho^\text{HS}_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be constant.
As such, ART coverage may also be modelled using a binomial logistic regression model
\begin{align}
A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
\end{align}
where the proportion of people receiving ART is $\rho^\text{HS}_i \alpha_i$.
Here we assume no travel between districts to receive treatment.

@@

Let $\y = (\y^\text{HS}, \y^\text{ANC}, \bm{A})$ be the complete response vector.
In this section, we would like to set-up notation for the joint model, helping to show that it isn't a latent Gaussian model, but is an extended latent Gaussian model.
When introducing LGM and ELGM below, we may refer back to this model as example.
For example, we can note that each of the components individually might be latent Gaussian (perhaps for the ART component it depends if you consider $\rho^\text{HS}_i$ to be fixed) but when combined they are no longer.

@@

<!-- ## Integrated nested Laplace approximation\label{sec:inla} -->

<!-- Consider a latent Gaussian model (LGM) of the form -->
<!-- \begin{alignat}{2} -->
<!-- &\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\ -->
<!-- &\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\ -->
<!-- &\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters} -->
<!-- \end{alignat} -->
<!-- where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$. -->
<!-- For such models, the joint posterior of $(\x, \btheta)$ is given by -->
<!-- \begin{equation} -->
<!--   p(\x, \btheta \, | \, \y) -->
<!--   \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right). \label{eq:lgmjointposterior} -->
<!-- \end{equation} -->
<!-- Rather than approximating this joint posterior, the INLA method proceeds by approximation of the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by -->
<!-- \begin{align} -->
<!--   p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\ -->
<!--   p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2} -->
<!-- \end{align} -->
<!-- An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$. -->
<!-- We discuss these two approximations in turn below. -->

<!-- ### Approximation (i) \label{sec:approxi} -->

<!-- The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}. -->
<!-- This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration nodes $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$. -->
<!-- The first step is to rewrite $p(\btheta \, | \, \y)$ as -->
<!-- \begin{equation} -->
<!--   p(\btheta \, | \, \y) -->
<!--   = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)} -->
<!--   \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost} -->
<!-- \end{equation} -->
<!-- Approximation (i) then uses a Gaussian approximation to the denominator given by -->
<!-- \begin{equation} -->
<!-- p(\x \, | \, \btheta, \y) \approx p_\text{G}(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \bm{\mu}(\btheta), {\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx} -->
<!-- \end{equation} -->
<!-- This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution, given by -->
<!-- \begin{equation} -->
<!-- p(\x \, | \, \btheta, \y) \propto -->
<!-- \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right), -->
<!-- \end{equation} -->
<!-- close to being Gaussian because $\y$ tends not to be strongly informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial]. -->
<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference? Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$. -->
<!-- As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}. -->
<!-- Taking $\x = \bm{\mu}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as -->
<!-- \begin{equation} -->
<!-- \tilde p_\text{LA}(\btheta \, | \, \y) \propto \frac{p(\y, \x, \btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \bm{\mu}(\btheta)} = \frac{p(\y, \bm{\mu}(\btheta), \btheta)}{\det({\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal} -->
<!-- \end{equation} -->
<!-- where the equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\bm{\mu}(\btheta)$. -->
<!-- We describe this as a Laplace approximation to the parameter posterior. -->

<!-- ### Approximation (ii) -->

<!-- We now move on to the approximation $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$. -->
<!-- Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ for the denominator in Section \ref{sec:approxi} above, a natural approach [@rue2007approximate] is to marginalise this distribution directly to obtain -->
<!-- \begin{equation} -->
<!-- \tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, {\mu}_i(\btheta), 1 / {q}_i(\btheta)), -->
<!-- \end{equation} -->
<!-- where the marginal mean ${\mu}_i(\btheta)$ and precision ${q}_i(\btheta)$ are recovered directly from the relevant entries of $\bm{\mu}(\btheta)$ and ${\bm{Q}}(\btheta)$ respectively. -->
<!-- However, although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode, so is generally not advised. -->
<!-- Instead, @rue2009approximate propose two methods, a Laplace approximation and a simplified version which is less computationally demanding. -->
<!-- The full Laplace approximation is -->
<!-- \begin{align} -->
<!--   p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
<!--   &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
<!--   &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\ -->
<!--   &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)} -->
<!--   = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation} -->
<!-- \end{align} -->
<!-- where -->
<!-- \begin{equation} -->
<!-- p_G(\x_{-i} \, | \, x_i, \btheta, \y) = \mathcal{N}(\x_{-i} \, | \, \bm{\mu}_{-i}(x_i, \btheta), {\bm{Q}}_{-i}(x_i, \btheta)), -->
<!-- \end{equation} -->
<!-- is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\bm{\mu}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\bm{\mu}_{-i}(x_i, \btheta)$ is not the same as $\bm{\mu}_{-i}(\btheta)$.} -->
<!-- The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well. -->
<!-- However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $i$, which is often computationally prohibitive. -->
<!-- Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by @rue2009approximate to reduce the computational cost: -->
<!-- \begin{enumerate} -->
<!-- \item -->
<!-- Avoiding having to find the mode via optimisation by using the approximation $\bm{\mu}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$ -->
<!-- \item -->
<!-- As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$ -->
<!-- \end{enumerate} -->

Consider a latent Gaussian model (LGM) of the form
\begin{alignat}{2}
&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \label{eq:data} \\
&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\
&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
\end{alignat}
where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
The joint posterior of $(\x, \btheta)$ is given by
\begin{equation}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right). \label{eq:lgmjointposterior}
\end{equation}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}

@@

Given a \textsf{C++} user template `model.cpp` for the unnormalised log posterior, we obtain posterior marginal approximations via the following algorithm:

@@

3. Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(x_i, \btheta, \y)$
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$.

4. Normalise $\tilde p_\text{LA}(x_i, \btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
$$
\tilde p_\text{AQ}(x_i \, | \, \btheta, \y) = \frac{\tilde p_\text{LA}(x_i, \btheta, \y)}{\tilde p_{\text{AQ}}(\btheta, \y)},
$$
where the normalising constant is calculated using nodes $z_x \in \mathcal{Q}(1, k_x)$ as
$$
\tilde p_{\text{AQ}}(\btheta, \y) = \sum_{z_x \in \mathcal{Q}(1, k_x)} \tilde p_\text{LA}(x_i(z_x), \btheta, \y) \omega_x(z_x),
$$
where the total number of nodes is $k_x$ and weights $\omega_x: z_x \in \mathcal{Q}(1, k_x) \to \mathbb{R}$.
As this quadrature is in one dimension, we can set $k_x = 5$ or higher.

5. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by
$$
\tilde p(x_i \, | \, \y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y) \tilde p_\text{AQ} (\btheta(\z_\theta) \, | \, \y) \omega_\theta(\z_\theta).
$$

@@

The \textsc{R} [@r] code used to implement the model and produce all results we describe is available at `github.com/athowes/elgm-inf`.
The inference method is available in versions 0.5.0. onwards of the `aghq` package.

@@

# References {#references .unnumbered}

>>>>

---
title: Integrated nested Laplace approximations for extended latent Gaussian models
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A
    sepNext: ","
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
  - name: Seth R.
    surname: Flaxman
    email: seth.flaxman@cs.ox.ac.uk
    label: e3
    addressLabel: C
    sepNext: ","
  - name: Jeffrey W.
    surname: Eaton
    email: jeffrey.eaton@imperial.ac.uk
    label: e4
    addressLabel: D
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
  - label: C
    name: Department of Computer Science, University of Oxford
    authorsLabels: e3
  - label: D
    name: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
    authorsLabels: e4
abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of interest, including HIV prevalence, HIV incidence and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. We propose a new inference method which combines the simplified integrated nested Laplace approximation approach of @wood2020simplified with adaptive Gauss-Hermite quadrature to enable fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method provides substantially more accurate inferences than the empirical Bayes Gaussian approximation approach used currently, and is comparable to Hamiltonian Monte Carlo with the No-U-Turn sampler. By extending the \texttt{aghq} \textsc{R} package we facilitate flexible and easy use of our method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - spatial statistics
  - small-area estimation
  - INLA
  - AGHQ
  - HIV epidemiology

@@

The complexity of the model presents a difficult Bayesian inference problem.
Any inferential strategy must be fast, as well as easy to run in production by country teams, ruling out prohibitively slow Markov chain Monte Carlo (MCMC) approaches.
Inference is currently conducted using an empirical Bayes approach, with a Gaussian approximation to the latent field, via the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb].
Owing to its speed and flexibility, `TMB` is gaining popularity spatial statistics [@osgoodzimmerman2021statistical].
Inference in `TMB` is based on optimisation of a \textsf{C++} template function, with the option available to use a Laplace approximation to integrate out any subset of the function arguments.
For the Naomi model, we use this option to integrate out the latent field parameters.
Taking inspiration from the AD Model Builder (ADMB) package [@fournier2012ad], `TMB` uses automatic differentiation [@baydin2017automatic] to calculate the derivatives required for numerical optimisation routines and the Laplace approximation.
Although this approach has favourable computational properties, we have found the inferences generated for the Naomi model to sometimes be inaccurate.

To obtain fast, accurate inferences for the Naomi model we develop a new inference methodology which combines the simplified integrated nested Laplace approximation (INLA) approach of @wood2020simplified with adaptive Gauss-Hermite quadrature (AGHQ).
INLA is an approach to approximate Bayesian inference based on nested Laplace approximations and numerical quadrature.
The central innovation of @rue2009approximate is a way to approximate accurate latent field posterior marginals without explicitly computing the full Laplace approximation for each element.
Simplified INLA [@wood2020simplified] extends INLA by relaxing the sparsity assumptions on the latent field required for this approximation to be accurate.
This extension facilitates inference for models like Naomi, which fall within the extended latent Gaussian model (ELGM) [@stringer2022fast] class, and were not previously amenable to inference with INLA.
ELGMs build on latent Gaussian models (LGMs) by allowing each element of the linear predictor to depend on any subset of elements from the latent field.
We combine simplified INLA with AGHQ, a quadrature rule based on the theory of polynomial interpolation which adapts to the integrand based on the Hessian at the mode.
Though no theory yet exists for the nested case, the first stochastic convergence results for adaptive quadrature rules were recently obtained by @bilodeau2021stochastic using AGHQ.
We implement our method as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
As `aghq` is designed to naturally interface with `TMB`, use of the method is easy when provided a \textsf{C++} user template for the log-posterior.

@@

Consider a country partitioned into $n$ areas indexed by $i$.
Suppose a simple random household survey of $m^\text{HS}_i$ people is conducted in each area, and $y^\text{HS}_i$ HIV positive cases are observed.
Then, cases may be modelled using a binomial logistic regression model
\begin{align*}
y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2),
\end{align*}
where $\rho^\text{HS}_i$ is the HIV prevalence, modelled on the logit scale by a Gaussian with mean $\beta_\phi \sim p(\beta_\phi)$ and standard deviation $\sigma_\phi \sim p(\sigma_\phi)$.

@@

Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
Suppose that of $m^\text{ANC}_i$ women attending ANC, $y^\text{ANC}_i$ are HIV positive.
Then an analogous binomial logistic regression model
\begin{align*}
y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
\end{align*}
may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
Reflecting the fact that prevalence in ANCs is related, but importantly different, to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence on the logit scale.
We use a Gaussian distribution with mean $\beta_b \sim p(\beta_b)$ and standard deviation $\sigma_b \sim p(\sigma_b)$ for these bias random effects.

@@

The number of people receiving treatment at district health facilities $A_i$ provides additional information about HIV prevalence.
Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa districts with low prevalence are likely to have fewer people recieving treatment.
ART coverage, defined to be the proportion of people living with HIV (PLHIV) currently on ART on district $i$, is given by $\alpha_i = A_i / \rho^\text{HS}_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be constant.
As such, ART coverage may also be modelled using a binomial logistic regression model
\begin{align*}
A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2),
\end{align*}
where the proportion of people receiving ART is $\rho^\text{HS}_i \alpha_i$.
We again use a Gaussian distribution with mean $\beta_\alpha \sim p(\beta_\alpha)$ and standard deviation $\sigma_\alpha \sim p(\sigma_\alpha)$.
Here we assume no travel between districts to receive treatment.

@@

The three components above may be combined to a single model as follows
\begin{align*}
y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
A_i &\sim \text{Bin}(N_i, \rho^\text{HS}_i \alpha_i), \\
\text{logit}(\rho^\text{HS}_i) &\sim \mathcal{N}(\beta_\phi, \sigma_\phi^2), \\
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho^\text{HS}_i) + b_i, \\
b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2), \\
\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
\end{align*}

@@

Consider a latent Gaussian model (LGM) of the form
\begin{alignat*}{2}
&\text{(Observations)}     &        \y &\sim p(\y \, | \, \x, \btheta), \\
&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \\
&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta),
\end{alignat*}
where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
The joint posterior of $(\x, \btheta)$ is given by
\begin{equation*}
  p(\x, \btheta \, | \, \y)
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right).
\end{equation*}
We consider approximations to the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
\begin{align}
  \tilde p(x_i \, | \, \y) \approx p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  \tilde p(\theta_j \, | \, \y) \approx p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}

@@

Given a \textsf{C++} user template `model.cpp` for the unnormalised log posterior, we obtain the posterior marginal approximations given in Equations \eqref{eq:inla1} and \eqref{eq:inla2} via the following algorithm:

@@

3. Obtain a nested approximation to the posterior marginal of the $i$th latent effect by
$$
\tilde p(x_i \, | \, \y) = \sum_{\z_\theta \in \mathcal{Q}(m, k_\theta)} \tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y) \tilde p_\text{AQ} (\btheta(\z_\theta) \, | \, \y) \omega_\theta(\z_\theta).
$$
The nodes and weights $\{\mathcal{Q}(m, k_\theta), \omega_\theta\}$ used to obtain $\tilde p_{\text{AQ}}(\y)$ are reused to perform integration with respect to the hyperparameters above.
For each of the $k_\theta^m$ values of $\btheta(\z_\theta)$ we obtain $\tilde p_\text{AQ}(x_i \, | \, \btheta(\z_\theta), \y)$ by setting $\btheta = \btheta(\z_\theta)$ in the following two steps.

a) Use a Laplace approximation to obtain the unnormalised $\tilde p_\text{LA}(x_i, \btheta, \y)$
$$
\tilde p_\text{LA}(x_i, \btheta, \y) = \frac{p(x_i, \x_{-i}, \btheta, \y)}{p_\text{G}(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \bm{\mu}_{-i}(x_i, \btheta)}
$$
where $\tilde p_\text{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \bm{\mu}_{-i}(x_i, \btheta), \bm{Q}_{-i}(x_i, \btheta)^{-1})$.

b) Normalise $\tilde p_\text{LA}(x_i, \btheta, \y)$ using adaptive Gauss-Hermite quadrature to obtain
$$
\tilde p_\text{AQ}(x_i \, | \, \btheta, \y) = \frac{\tilde p_\text{LA}(x_i, \btheta, \y)}{\tilde p_{\text{AQ}}(\btheta, \y)},
$$
where the normalising constant is calculated using nodes $z_x \in \mathcal{Q}(1, k_x)$ as
$$
\tilde p_{\text{AQ}}(\btheta, \y) = \sum_{z_x \in \mathcal{Q}(1, k_x)} \tilde p_\text{LA}(x_i(z_x), \btheta, \y) \omega_x(z_x),
$$
where the total number of nodes is $k_x$ and weights $\omega_x: z_x \in \mathcal{Q}(1, k_x) \to \mathbb{R}$.
As this quadrature is in one dimension, we can set $k_x = 5$ or higher.

@@

* The \textsc{R} [@r] code used to produce all results we describe is available at `github.com/athowes/elgm-inf`. The inference method is available in versions 0.5.0. onwards of the `aghq` package
* Using the `TMB` template, we fit the model using four inferential approaches: (1) empirical Bayes combined with a Gaussian approximation, (2) AGHQ combined with a Gaussian approximation, (3) AGHQ combined with a Laplace approximation and (4) the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS). We treat results from NUTS as the gold-standard
* We used the Kolmogorov-Smirnov test for the maximum difference between two empirical cumulative distribution functions to compare posterior marginal distributions
* We performed posterior predictive checks to assess the coverage of our estimates via the uniformity of the data within each posterior marginal distribution

<!-- \begin{table}[] -->
<!-- \small -->
<!-- \begin{tabularx}{\textwidth}{llll} -->
<!-- \toprule -->
<!-- Inference method & KS & MMD & PSIS \\ -->
<!-- \midrule -->
<!-- Empirical Bayes, Gaussian & 0 & 0 & 0 \\ -->
<!-- AGHQ, Gaussian & 0 & 0 & 0 \\ -->
<!-- AGHQ, Laplace & 0 & 0 & 0 \\ -->
<!-- NUTS & 0 & 0 & 0 \\ -->
<!-- \bottomrule -->
<!-- \end{tabularx} -->
<!-- \label{tab:results} -->
<!-- \small -->
<!-- \end{table} -->

@@
